|

### 2023-01-11 paper.tex
Wed Jan 11 19:00:10 CET 2023: update-remote.sh

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{float}
\usepackage[numbers]{natbib}
\let\origfigure\figure
\let\endorigfigure\endfigure
\renewenvironment{figure}[1][2] {
    \expandafter\origfigure\expandafter[H]
} {
    \endorigfigure
}
\usepackage{float}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

Johannes Gutenberg University Mainz}
\author{}
\date{\vspace{-2.5em}Date of Submission: 2022-12-29}

@@

Denoising Autoencoder Genetic Programming (DAE-GP) is a novel variation
of an genetic programming based Estimation of Distribution Algorithm
(EDA-GP) that uses a denoising autoencoders long short-term memory
network (DAE-LSTMs) as a probabilistic model to sample new candidate
solutions {[}18{]}.

DAE-LSTMs are recurrent neural networks (RNNs) that can be trained in an
unsupervised learning environment to minimize a reconstruction error for
encoding input data into a compressed representation and subsequently
decoding the compressed representation back to the input dimension. In
DAE-GP, DAE-LSTMs are trained with a subset of high-fitness solutions
selected from a parent population with the aim to capture their
promising qualities. The resulting model is then used to sample new
offspring solutions by propagating partially mutated solutions from the
parent population through the DAE-LSTMS {[}18{]}. In previous work
DAE-GP has been shown to outperform GP for both a generalized version of
the royal tree problem {[}18{]} as well as for a real-world symbolic
regression problem {[}17{]}.

The DAE-GP algorithm first described by {[}18{]} trains a DAE-LSTMs for
each generation \(g\) of the search from scratch. {[}17{]} and {[}16{]}
suggests the incorporation of a pre-training strategy into the
evolutionary search as a possible way of improving the model quality of
DAE-GP. The key idea is to pre-train an initial DAE-LSTM on a large
population of candidate solutions and to use the weights of this initial
model in each generation of the search as a starting point. This thesis
studies the influence of using pre-trained DAE-LSTMs in DAE-GP for
symbolic regression, especially looking at the influence in
generalization behavior and the overall quality of solutions found by
DAE-GP. The aim of this study is to answer the question if a
pre-training strategy can help to improve generalization in DAE-GP and
if so, analyze how this improved generalization ability benefits the
overall performance of DAE-GP.

@@

DAE-GP is an EDA-GP algorithm that uses DAE-LSTM networks as a
probabilistic model to sample new offspring solutions {[}18{]}.

@@

As a variant of GP, DAE-GP can be classified as a meta-heuristic that
belongs to the field of evolutionary computation (EC). EC based
meta-heuristics are optimization methods that simulate the process of
Darwinian evolution to search for high quality solutions by applying
selection and variation operators to a population of candidate
solutions. Examples of EC include genetic algorithms (GA), evolutionary
strategies (ES) and GP. In EC, the quality of a solution is commonly
measured as fitness and the time steps of the search are called
generations. Another important concept in EC is the distinction between
genotypes and phenotypes of solutions, the genotype contains the
information that is necessary to construct the phenotype, the outer
appearance of a particular solution on which we measure the overall
quality of solutions. The representation of a solution is therefore
defined by the mapping of genotypes to phenotypes {[}12{]}. Genetic
operators, such as mutation or recombination are usually applied to the
genotype of solutions.

@@

GP follows the same basic evolutionary principle of EC but searches for
more general, hierarchical computer programs of dynamically varying size
and shape {[}8{]}. The computer programs that are at the center of the
evolutionary search in GP are commonly represented by tree structures at
the level of their phenotype {[}12{]}. Since GP searches for high
fitness computer programs that produce a desired output for some input,
it can be applied to various different problem domains such as symbolic
regression, automatic programming, or evolving game-playing strategies
{[}8{]}. An important quality of GP is the ability to search for
solutions of variable length and structure. GP is an especially useful
meta heuristic for problems where no a priori knowledge about the final
form of good solutions is available. GPs ability to optimize solutions
for their structure as well as for their parameters led to it being one
of the most prevalent methods used symbolic regression {[}10{]}.

@@

The aim of Estimation of distribution algorithms (EDA) is to replace the
standard variation operators used in GA by building probabilistic models
that can capture complex dependencies between different decision
variables of an optimization problem {[}12{]}. EDAs use this
probabilistic model to sample new offspring solutions inside an
evolutionary search to replace crossover and/or mutation operators.

@@

One possible way of model building in EDA proposed by {[}11{]} is to use
denoising autoencoders (DAEs) as generative models to capture complex
probability distributions of decision variables.

DAE, a variation of the autoencoders (AE), are a widely used type of
neural networks in the domain of unsupervized machine learning that maps
\(n\) input variables to \(n\) output variables using a hidden
representation.

AE were introduced by {[}6{]} to compress high-dimensional data into
lower-dimensions. An AE consists of two different subunits:

@@

The AE is trained to reduce the reconstruction error between input and
output data, after the training procedure is finished the network is
able to reduce the dimensionality of input data to a compressed
representation{[}6{]}.

DAE was first introduced by {[}14{]} as an improved AE with the ability
to learn new representations of data that is especially robust to
partially corrupted input data. DAE modifies the AE by using partially
corrupted input data for the AE and training it to reconstruct the
uncorrupted, original version of the input data.

Since the hidden representation of DAE captures the dependency structure
of the input data it can therefore also be used to generate new
solutions in the context of GAs {[}11{]}.

DAE-GP builds upon the concept of using DAEs in EDAs described by
{[}11{]} and transfers the concept to the domain of GP. The mutation and
crossover operators of standard GP are replaced by sampling new
solutions from a probabilistic model that is build by training a DAEs
long short-term memory (LSTM) network on a subset of high fitness
solutions from the current population {[}18{]}.

LSTMs are a variant of recurrent neural networks first introduced by
{[}7{]} that can store learned information over en extended time period
while avoiding the problem of vanishing and/or exploding gradients.
Since DAE-GP encodes candidate solutions as linear strings in prefix
notation, the DAE in DAE-GP uses LSTMs for both encoding and decoding
where the total amount of time steps \(T\) is equal to sum of the length
of the input solution and the output solution {[}18{]}.

@@

Pre-Training describes the concept of initially training a neural
network on a large dataset before using it to solve a more specific
task. It is a commonly used strategy in deep architectures that has been
shown to improve both the optimization process itself as well as the
generalization behavior if compared to the standard approach of using
randomly initialized parameters {[}3{]}.

The strategy of pre-training artificial neural networks is based on the
idea of transfer learning that aims at retaining previously learned
knowledge from one task to re-use it for another task {[}5{]} {[}9{]}.
Transfer learning traditionally follows a two phase approach:

@@

where source task and target task are usually similar but may differ in
their feature space and the distribution of training data {[}9{]}.

@@

The pre-training strategy used in this thesis is applied to the DAE-LSTM
model \(M_g\) that are used in each DAE-GP generation \(g\) for
\(g\in{1,...,g_{max}}\). Instead of initializing and optimizing each
model from scratch as done in previous work (e.g. {[}18{]} {[}17{]}), a
separate DAE-LSTM network \(\hat{M}\) will be trained on a large
population of randomly initialized solutions.\\
The optimal parameters obtained after finishing the training procedure
of \(\hat{M}\) are then used as the starting parameters for each
following DAE-LSTM model \(M_g\) for \(g\in{1,...,g_{max}}\).

@@

The probably most obvious motivation for using a pre-training strategy
in DAE-GP is based on the fact, that initializing \(M_g\) with
pre-trained weights from \(\hat{M}\) is likely to reduce the amount of
training epochs that are necessary until the point of training error
convergence is reached.

Early research on pre-training for DAE by {[}3{]} comes to the
conclusion that besides adding robustness to models the strategy also
results in improved generalization and better performing models. The
authors describe that unsupervised pre-training behaves like a regulizer
on DAE networks, the mean test error and it's variance are reduced for
DAE networks that are initialized from pre-trained weights if compared
to the same architectures that use random initialization.

Another important finding by {[}3{]} is that the positive effect of
pre-training described above is dependent on both the depth of the
network as well as the size of layers - while increasingly larger
networks benefit increasingly more from pre-training the final
performance of small architectures tends to be worse with pre-training
than with randomized weight initialization. The authors find evidence
that pre-training DAE is especially usefull for optimizing the
parameters of the lower layers of the network.

Another motivation for introducing pre-training into DAE-GP is the
prevalence of sampling error for small GP populations sizes. {[}13{]}
finds that sampling error, non-systematic errors that are caused by
observing only a small subset of the statistical population, is a severe
problem in the domain of GP since the initial population may not be a
representative sample of all possible solutions. {[}13{]} also
introduces a method for calculating optimal population sizes to minimize
the presence of sampling error. Since DAE-LSTM of DAE-GP learns the
properties of it's training population and reuses the acquired knowledge
in the sampling procedure, using the parameters obtained from training
\(\hat{M}\) on a sufficiently large training population might reduce the
need for large population sizes in the following generations of the
search if \(\hat{M}\) already implicitly captured the properties of a
representative sample of solutions. If this mechanism can be applied
successfully it would benefit the performance of DAE-GP by increasing
the population diversity (resulting in better solution quality) as well
as by reducing the need for large population sizes which require
computational resources.

@@

The DAE-GP algorithm, first described by {[}18{]}, is summarized by
figure 1.

\begin{figure}
\centering
\includegraphics[width=2.08333in,height=8.33333in]{./img/flowcharts/dae-gp.png}
\caption{Flowchart - Regular DAE-GP}
\end{figure}

The pre-training strategy implemented is visualized by figure 2.

\begin{figure}
\centering
\includegraphics[width=2.08333in,height=8.33333in]{./img/flowcharts/pt-dae-gp.png}
\caption{Flowchart - Pre-Trained DAE-GP}
\end{figure}

One difficulty in the implementation of a pre-training strategy into
DAE-GP has been the detemination of the number of hidden neurons in side
the DAE-LSTMs hidden layers. The original description of DAE-GP {[}18{]}
used a strategy where the number of hidden neurons for each hidden layer
is dynamically set per generation to the maximum individual size inside
the current population. For a pre-training implementation, this strategy
can not be easily adapted since it leads to a changing number of neurons
at each generation resulting in different dimensions of the DAE-LSTM. To
allow for the sharing of weights and biases from the pre-trained model
\(\hat{M}\) to each \(M_g\) for \(g\in{1,...,g_{max}}\) I experimented
with two different strategies:

@@

The framework that was used to conduct all experiments of this thesis
was provided by the supervisor David Wittenberg and uses the python
programming language in conjunction with the baseline libraries
\texttt{Keras} {[}1{]} for deep neural networks and \texttt{deap}
{[}4{]} for evolutionary computation.

@@

To test pre-training in DAE-GP this thesis focuses on the domain of
real-world symbolic regression problems. Symbolic Regression problems
have been one of the first GP applications {[}8{]} and are an actively
studied and highly relevant research area. The goal in symbolic
regression is to find a mathematical model for a given set of data
points {[}10{]}, in real-world symbolic regression these data points are
sourced from real-world observations which in contrast to synthetic
symbolic regression problem are more likely to contain random noise and
bias. Another important challenge in solving real-world symbolic
regression problems is the ability for a given model to generalize, we
want the final model to show high accuracy in predicting outcomes for
previously unseen cases.

The main experiments conducted in this thesis uses the NASA Airfoil
Self-Noise Data Set which is part of the UCI machine learning repository
{[}2{]}. The dataset consists of 5 input variables and 1 output variable
that are listed in table 1.

@@

{[}3{]}: Pre-Training effect especially usefull for lower-level layers
-\textgreater{} Only adapt weights from low layers or embedd the
pre-trained model?

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{float}
\usepackage[numbers]{natbib}
\let\origfigure\figure
\let\endorigfigure\endfigure
\renewenvironment{figure}[1][2] {
    \expandafter\origfigure\expandafter[H]
} {
    \endorigfigure
}
\usepackage{float}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

Johannes Gutenberg University Mainz}
\author{}
\date{\vspace{-2.5em}Date of Submission: 2023-01-11}

@@

Denoising Autoencoder Genetic Programming (DAE-GP) is a novel variation
of an genetic programming based Estimation of Distribution Algorithm
(EDA-GP) that uses a denoising autoencoders long short-term memory
network (DAE-LSTMs) as a probabilistic model to sample new candidate
solutions {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{18}{]}.

DAE-LSTMs are recurrent neural networks (RNNs) that can be trained in an
unsupervised learning environment to minimize a reconstruction error for
encoding input data into a compressed representation and subsequently
decoding the compressed representation back to the input dimension. In
DAE-GP, DAE-LSTMs are trained with a subset of high-fitness solutions
selected from a parent population with the aim to capture their
promising qualities. The resulting model is then used to sample new
offspring solutions by propagating partially mutated solutions from the
parent population through the DAE-LSTMS
{[}\protect\hyperlink{ref-dae-gp_2020_rtree}{18}{]}. In previous work
DAE-GP has been shown to outperform GP for both a generalized version of
the royal tree problem
{[}\protect\hyperlink{ref-dae-gp_2020_rtree}{18}{]} as well as for a
real-world symbolic regression problem
{[}\protect\hyperlink{ref-dae-gp_2022_symreg}{17}{]}.

The DAE-GP algorithm first described by
{[}\protect\hyperlink{ref-dae-gp_2020_rtree}{18}{]} trains a DAE-LSTMs
for each generation \(g\) of the search from scratch.
{[}\protect\hyperlink{ref-dae-gp_2022_symreg}{17}{]} and
{[}\protect\hyperlink{ref-daegp_explore_exploit}{16}{]} suggests the
incorporation of a pre-training strategy into the evolutionary search as
a possible way of improving the model quality of DAE-GP. The key idea is
to pre-train an initial DAE-LSTM on a large population of candidate
solutions and to use the pre-trained parameters of this initial model in
each generation of the search as starting parameters for the current
generation DAE-LSTM. This thesis studies the influence of using
pre-trained DAE-LSTMs in DAE-GP for symbolic regression, especially
looking at the influence in generalization behavior and the overall
quality of solutions found by DAE-GP. The aim of this study is to answer
the question if a pre-training strategy can help to improve
generalization in DAE-GP and if so, analyze how this improved
generalization ability benefits the overall performance of DAE-GP.

@@

DAE-GP is an EDA-GP algorithm that uses DAE-LSTM networks as a
probabilistic model to sample new offspring solutions
{[}\protect\hyperlink{ref-dae-gp_2020_rtree}{18}{]}.

@@

As a variant of GP, DAE-GP can be classified as a meta-heuristic that
belongs to the field of evolutionary computation (EC). EC based
meta-heuristics are optimization methods that simulate the process of
Darwinian evolution to search for high quality solutions by applying
selection and variation operators to a population of candidate
solutions. Examples of EC include genetic algorithms (GA), evolutionary
strategies (ES) and GP. In EC, the quality of a solution is commonly
measured as fitness and the time steps of the search are called
generations. Another important concept in EC is the distinction between
genotypes and phenotypes of solutions, the genotype contains the
information that is necessary to construct the phenotype, the outer
appearance of a particular solution on which we measure the overall
quality of solutions. The representation of a solution is therefore
defined by the mapping of genotypes to phenotypes
{[}\protect\hyperlink{ref-design_of_modern_heuristics}{12}{]}. Genetic
operators, such as mutation or recombination are usually applied to the
genotype of solutions.

@@

GP follows the same basic evolutionary principle of EC but searches for
more general, hierarchical computer programs of dynamically varying size
and shape {[}\protect\hyperlink{ref-Koza1993GeneticP}{8}{]}. The
computer programs that are at the center of the evolutionary search in
GP are commonly represented by tree structures at the level of their
phenotype {[}\protect\hyperlink{ref-design_of_modern_heuristics}{12}{]}.
Since GP searches for high fitness computer programs that produce a
desired output for some input, it can be applied to various different
problem domains such as symbolic regression, automatic programming, or
evolving game-playing strategies
{[}\protect\hyperlink{ref-Koza1993GeneticP}{8}{]}. An important quality
of GP is the ability to search for solutions of variable length and
structure. GP is an especially useful meta heuristic for problems where
no a priori knowledge about the final form of good solutions is
available. GPs ability to optimize solutions for their structure as well
as for their parameters led to it being one of the most prevalent
methods used for symbolic regression
{[}\protect\hyperlink{ref-10.1007ux2f978-3-540-24621-3_22}{10}{]}.

@@

The aim of Estimation of distribution algorithms (EDA) is to replace the
standard variation operators used in GA by building probabilistic models
that can capture complex dependencies between different decision
variables of an optimization problem
{[}\protect\hyperlink{ref-design_of_modern_heuristics}{12}{]}. EDAs use
this probabilistic model to sample new offspring solutions inside an
evolutionary search to replace crossover and/or mutation operators.

@@

One possible way of model building in EDA proposed by
{[}\protect\hyperlink{ref-harmless_overfitting_eda}{11}{]} is to use
denoising autoencoders (DAEs) as generative models to capture complex
probability distributions of decision variables. DAE, a variation of the
autoencoders (AE), are a widely used type of neural networks in the
domain of unsupervized machine learning that maps \(n\) input variables
to \(n\) output variables using a hidden representation.

AE were introduced by {[}\protect\hyperlink{ref-ae_orig}{6}{]} to
compress high-dimensional data into lower-dimensions. An AE consists of
two different subunits:

@@

The AE is trained to reduce the reconstruction error between input and
output data, after the training procedure is finished the network is
able to reduce the dimensionality of input data to a compressed
representation{[}\protect\hyperlink{ref-ae_orig}{6}{]}.

DAE was first introduced by
{[}\protect\hyperlink{ref-10.1145ux2f1390156.1390294}{14}{]} as an
improved AE with the ability to learn new representations of data that
is especially robust to partially corrupted input data. DAE modifies the
AE by using partially corrupted input data for the AE and training it to
reconstruct the uncorrupted, original version of the input data.

Since the hidden representation of DAE captures the dependency structure
of the input data it can therefore also be used to generate new
solutions in the context of GAs
{[}\protect\hyperlink{ref-harmless_overfitting_eda}{11}{]}.

DAE-GP builds upon the concept of using DAEs in EDAs described by
{[}\protect\hyperlink{ref-harmless_overfitting_eda}{11}{]} and transfers
the concept to the domain of GP. The mutation and crossover operators of
standard GP are replaced by sampling new solutions from a probabilistic
model that is build by training a DAEs long short-term memory (LSTM)
network on a subset of high fitness solutions from the current
population {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{18}{]}.

LSTMs are a variant of recurrent neural networks first introduced by
{[}\protect\hyperlink{ref-lstm_orig}{7}{]} that can store learned
information over en extended time period while avoiding the problem of
vanishing and/or exploding gradients. Since DAE-GP encodes candidate
solutions as linear strings in prefix notation, the DAE in DAE-GP uses
LSTMs for both encoding and decoding where the total amount of time
steps \(T\) is equal to sum of the length of the input solution and the
output solution {[}\protect\hyperlink{ref-dae-gp_2020_rtree}{18}{]}.

@@

Pre-Training describes the concept of initially training a neural
network on a large dataset before using it to solve a more specific
task. It is a commonly used strategy in deep architectures that has been
shown to improve both the optimization process itself as well as the
generalization behavior if compared to the standard approach of using
randomly initialized parameters
{[}\protect\hyperlink{ref-pmlr-v5-erhan09a}{3}{]}.

The strategy of pre-training artificial neural networks is based on the
idea of transfer learning that aims at retaining previously learned
knowledge from one task to re-use it for another task
{[}\protect\hyperlink{ref-HAN2021225}{5}{]}
{[}\protect\hyperlink{ref-survey_transfer_learning}{9}{]}. Transfer
learning traditionally follows a two phase approach:

@@

where source task and target task are usually similar but may differ in
their feature space and the distribution of training data
{[}\protect\hyperlink{ref-survey_transfer_learning}{9}{]}.

@@

The pre-training strategy used in this thesis is applied to the DAE-LSTM
model \(M_g\) that are used in each DAE-GP generation \(g\) for
\(g\in{1,...,g_{max}}\). Instead of initializing and optimizing each
model from scratch as done in previous work (e.g.
{[}\protect\hyperlink{ref-dae-gp_2020_rtree}{18}{]}
{[}\protect\hyperlink{ref-dae-gp_2022_symreg}{17}{]}), a separate
DAE-LSTM network \(\hat{M}\) will be trained on a large population of
randomly initialized solutions.\\
The trainable parameters \(\theta_{\hat{M}}\) obtained after finishing
the training procedure of \(\hat{M}\) are then used as the starting
parameters for each following DAE-LSTM model \(M_g\) for
\(g\in{1,...,g_{max}}\).

@@

The probably most obvious motivation for using a pre-training strategy
in DAE-GP is based on the fact, that initializing \(M_g\) with
pre-trained weights \(\theta_{\hat{M}}\) is likely to reduce the amount
of training epochs that are necessary until the point of training error
convergence is reached.

Early research on pre-training for DAE by
{[}\protect\hyperlink{ref-pmlr-v5-erhan09a}{3}{]} comes to the
conclusion that besides adding robustness to models the strategy also
results in improved generalization and better performing models. The
authors describe that unsupervised pre-training behaves like a regulizer
on DAE networks, the mean test error and it's variance are reduced for
DAE networks that are initialized from pre-trained weights if compared
to the same architectures that use random initialization.

Another important finding by
{[}\protect\hyperlink{ref-pmlr-v5-erhan09a}{3}{]} is that the positive
effect of pre-training described above is dependent on both the depth of
the network as well as the size of layers - while increasingly larger
networks benefit increasingly more from pre-training the final
performance of small architectures tends to be worse with pre-training
than with randomized weight initialization. The authors find evidence
that pre-training DAE is especially usefull for optimizing the
parameters of the lower layers of the network.

Another motivation for introducing pre-training into DAE-GP is the
prevalence of sampling error for small GP populations sizes.
{[}\protect\hyperlink{ref-sampling_err_gp}{13}{]} finds that sampling
error, non-systematic errors that are caused by observing only a small
subset of the statistical population, is a severe problem in the domain
of GP since the initial population may not be a representative sample of
all possible solutions.
{[}\protect\hyperlink{ref-sampling_err_gp}{13}{]} also introduces a
method for calculating optimal population sizes to minimize the presence
of sampling error. Since DAE-LSTM of DAE-GP learns the properties of
it's training population and reuses the acquired knowledge in the
sampling procedure, using the parameters obtained from training
\(\hat{M}\) on a sufficiently large training population might reduce the
need for large population sizes in the following generations of the
search if \(\hat{M}\) already implicitly captured the properties of a
representative sample of solutions. If this mechanism can be applied
successfully it would benefit the performance of DAE-GP by increasing
the population diversity (resulting in better solution quality) as well
as by reducing the need for large population sizes which require
computational resources.

@@

The DAE-GP algorithm, first described by
{[}\protect\hyperlink{ref-dae-gp_2020_rtree}{18}{]}, is summarized as a
flowchart in figure §. After setting the generation counter \(g\) to
\(0\) and creating the initial population of solutions \(P_0\), DAE-GP
enters the main loop that checks if a termination criteria is satisfied
(i.e.~maximum number of generations or fitness evaluations is reached).
For each generation \(g\) the fitness of all new individuals in \(P_g\)
is evaluated. During model building, DAE-GP selects a subset \(X_g\) of
the current population that is used as training data for the DAE-LSTM
model \(M_g\) which is trained to learn the properties of \(X_g\). The
trained model \(M_g\) is then used for model sampling, partially mutated
solutions from \(P_G\) are propagated through \(M_g\) to create a new
population \(P_{g+1}\). This process is repeated until the termination
criteria is met and DAE-GP returns the highest fitness solutions.

\begin{figure}
\centering
\includegraphics{./img/flowcharts/dae-gp.png}
\caption{Flowchart - Regular DAE-GP}
\end{figure}

The pre-training strategy implemented for all experiments in this thesis
is visualized as another flowchart in figure §. The main difference to
regular DAE-GP is the inclusion of an initial pre-training phase where a
separate population \(\hat{P}\) is first initialized and then randomly
split by half into \(\hat{P}_{train}\) and \(\hat{P}_{test}\). A
DAE-LSTM model \(\hat{M}\) is then trained to learn the properties of
\(\hat{P}_{train}\) using an early stopping training mode where we stop
training as soon as the validation error for \(\hat{P}_{test}\)
converges. After the training for \(\hat{M}\) is stopped, the current
state of \(\hat{M}\) is frozen and the optimized trainable parameters
\(\theta_{\hat{M}}\) are saved before terminating the pre-training phase
to start the DAE-GP phase. During the DAE-GP phase, the only difference
to the traditional DAE-GP algorithm described in figure § is during
model building: We initialize \(M_g\) with \(\theta_{\hat{M}}\) before
training it to learn the properties of \(X_g\).

\begin{figure}
\centering
\includegraphics{./img/flowcharts/pt-dae-gp.png}
\caption{Flowchart - Pre-Trained DAE-GP}
\end{figure}

One difficulty in the implementation of a pre-training strategy into
DAE-GP has been the detemination of the number of hidden neurons in side
the DAE-LSTMs hidden layers. The original description of DAE-GP
{[}\protect\hyperlink{ref-dae-gp_2020_rtree}{18}{]} used a strategy
where the number of hidden neurons for each hidden layer is dynamically
set per generation to the maximum individual size inside the current
population. For a pre-training implementation, this strategy can not be
easily adapted since it leads to a changing number of neurons at each
generation resulting in different dimensions of the DAE-LSTM. To allow
for the sharing of weights and biases from the pre-trained model
\(\hat{M}\) to each \(M_g\) for \(g\in{1,...,g_{max}}\) I experimented
with two different strategies:

@@

The framework that was used to conduct all experiments of this thesis
was provided by the supervisor David Wittenberg and uses the python
programming language in conjunction with the baseline libraries
\texttt{Keras} {[}\protect\hyperlink{ref-chollet2015keras}{1}{]} for
deep neural networks and \texttt{deap}
{[}\protect\hyperlink{ref-DEAP_JMLR2012}{4}{]} for evolutionary
computation.

@@

To test pre-training in DAE-GP this thesis focuses on the domain of
real-world symbolic regression problems. Symbolic Regression problems
have been one of the first GP applications
{[}\protect\hyperlink{ref-Koza1993GeneticP}{8}{]} and are an actively
studied and highly relevant research area. The goal in symbolic
regression is to find a mathematical model for a given set of data
points
{[}\protect\hyperlink{ref-10.1007ux2f978-3-540-24621-3_22}{10}{]}, in
real-world symbolic regression these data points are sourced from
real-world observations which in contrast to synthetic symbolic
regression problem are more likely to contain random noise and bias.
Another important challenge in solving real-world symbolic regression
problems is the ability for a given model to generalize, we want the
final model to show high accuracy in predicting outcomes for previously
unseen cases.

The main experiments conducted in this thesis uses the NASA Airfoil
Self-Noise Data Set which is part of the UCI machine learning repository
{[}\protect\hyperlink{ref-machine_learning_repo}{2}{]}. The dataset
consists of 5 input variables and 1 output variable that are listed in
table 1.

@@

{[}\protect\hyperlink{ref-pmlr-v5-erhan09a}{3}{]}: Pre-Training effect
especially usefull for lower-level layers -\textgreater{} Only adapt
weights from low layers or embedd the pre-trained model?
